### 2023-11-14 differentiating_pHm.Rmd
differentiating pHm(): why pK1 is unimportant
I finally found a visual explanation of why pK1 cannot be reliably
estimated from real-world experimental data. Derivaitves very clearly
show that for HEPES, pK1 has a visible impact on the experimental data
only at very low pH values.

 * definitions.R: contains common functions. We may later move its
   contents inside the main .Rmd file.
 * differentiating_pHm.Rmd: the explanation itself.

```{r}
library(nlsr)
pK1 <- -1; pK2 <- 3; pK3 <- 7.55 # literature values
HEPTOT <- 0.05
kw <- 1e-14
source('definitions.R')
```

Differentiating the synthetic $\mathrm{pH}$ values will let us know how
every parameter impacts the value of the function. Thankfully, `nlsr`
already has the necessary tools for that, accessible from plain **R**:

```{r}
SID <- seq(-5e-2,4e-2,length=1000)
e <- quote(pHm(SID, HEPTOT, pK1, pK2, pK3))
pH <- numericDerivR(e, paste0('pK', 1:3), central = TRUE)
plot(SID, pH, type = 'l')
```

One minor detail: the values are different enough in their orders of
magnitude to warrant the use of logarithmic scale. There will be very
small negative numbers which we'll replace with `NA` to avoid warnings.

```{r}
g <- attr(pH, 'gradient')
g[g <= 0 & abs(g) < 1e-9] <- NA
par(mar = c(5, 6, 4, 2) + .1)
matplot(SID, g, type = 'l', lty = 1, col = 2:4, log = 'y', ylab = quote(
	frac(partialdiff*'pH', partialdiff*pK[j])
))
legend('bottomright', lty = 1, col = 2:4, legend = sapply(1:3,
	function(j) bquote(pK[.(j)])
))
```

This shows very clearly that the contribution of $\mathrm{p}K_1$ is very
minor and is only noticeable at significantly negative $\mathrm{SID}$,
which imply low $\mathrm{pH}$ values, which cannot be reliably measured.

### 2023-11-14 definitions.R

@@ -1,0 +1,17 @@
+# The charge balance function for a weak acid with three given pKs.  The
+# root of this function gives the equilibrium value of [H] at the given
+# conditions.
+HEPESFUNC <- function(H,SID,HEPTOT,pK1,pK2,pK3) {
+	XX <- (H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
+	IV <- HEPTOT/(1+XX)
+	I <- IV*H^3/(10^-pK3*10^-pK2*10^-pK1)
+	II <- IV*H^2/(10^-pK3*10^-pK2)
+	III <- IV*H/10^-pK3
+	H - kw/H + SID + I*2 + II - IV
+}
+
+# Return a vector of equilibrium pHs for the given SIDs and other
+# constant parameters.
+pHm <- Vectorize(function(SID, HEPTOT, pK1, pK2, pK3)
+  -log10(uniroot(HEPESFUNC,c(1e-20,1),tol=1e-20,maxiter=1e4,
+                 HEPTOT=HEPTOT,SID = SID, pK1=pK1,pK2=pK2,pK3=pK3)$root))

### 2023-11-14 manuscript.Rmd
Import the manuscript
This is a mostly clean conversion including all formulas and figures.
The WWW references are not fully satisfactory yet, but that could be
amended later. Currently, bookdown features (cross-references) are used.

---
title: Modeling HEPES buffer with a charge-balance approach
author: Troels Ring, Søren Risom Kristensen, Ivan Krylov, John C Nash.
bibliography: manuscript.bib
---

```{r, echo=FALSE}
# NOTE: bookdown features (e.g. cross-references) used below
# This requires the use of a bookdown output format, e.g.
# bookdown::pdf_document2 or bookdown::html_document2
```

Abstract
========

Introduction
============

It has been shown [@Ring2022] that $\mathrm{pH}$ in arbitrary fluids can be
understood based on the three fundamental principles of
electroneutrality, conservation of mass, and rules of dissociation as
devised from physical chemistry. It has also been shown [@Ring2022] that the
charge-balance approach allows definition of the $\mathrm{pH}$-dependent
buffer-capacity, and provides a definition of acid content as titratable
acidity, based on previous work by Kildeberg [@Kildeberg1983]. Recently
it has been reiterated [@Ring2023-charge] that it is absolutely necessary for this
modeling approach to take into account the strong ions, which are those
assumed to be completely dissociated at any $\mathrm{pH}$. In contrast to the
physico-chemical model of Stewart [@Stewart1978], the charge-balance
model (CBM, henceforth) has no separation between independent and
dependent variables. Under the CBM all that can be known about the
system is gathered in the expression of charge in an explicitly stated
chemical system. Consequently, if, as an example, $\mathrm{pH}$ is known and charge
is a known deterministic function of the chemical specification, inverse
inferences on unknown variables can, at least sometimes, be obtained by
root finding, since we know that the charge must be zero. In simple
systems, such as a single monovalent weak acid and strong ions, a single
unknown variable is explicitly estimable [@Ring2022]. The question
focused here is if CBM is rich enough to allow inverse inference in more
complex systems such as with multiprotic buffers. This question arose
under experiments with HEPES buffered sulfate containing fluids
[@Ring2023-acid] in which HEPES was modeled as a monovalent buffer with
pKa 7.5 [@Ring2010] since it was found that estimated $\mathrm{pH}$ was much too
low when observed $\mathrm{pH}$ was less than about 6. HEPES is a zwitterion with
formula 2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic acid and
structure as shown in \@ref(fig:hepes).

```{r hepes, fig.cap = '2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic acid', echo = FALSE}
knitr::include_graphics('HEPES.svg', auto_pdf = TRUE)
```

Wikipedia informs [@WP:HEPES] that HEPES has two pK values, one at 3 and
one at 7.5, but the reference given [@Johnson2014] in Wikipedia states
that "3" comes from the sulfonate group and 7.5 from one of the
piperazine nitrogens.  At $20^\circ\mathrm{C}$ Good et al [@Good1966]
has pKa 3 and 7.55 without detailing which groups are at work, while
another study [@AlGhobashy2014] mentions that 3 and 7.5 both belongs to
piperazine amines and informs that the sulfonate group is dissociated
over "almost the whole $\mathrm{pH}$ range".  Furthermore, is has been stated that
the nitrogen at the $\mathrm{OH}$ end of HEPES is protonated in methanol
and the nitrogen in the sulfonate end is protonated in water
[@Sledz2010].

Hence it was not clear how best to model the $\mathrm{pH}$-dependent
charge on HEPES to estimate $\mathrm{pH}$ based on known chemistry.
Furthermore, it was desired to make clear if the $\mathrm{p}K$ values
for HEPES could be elucidated based on measured $\mathrm{pH}$. Herein,
these questions will be approached by first developing a model for the
mean charge on HEPES as a function of $\mathrm{pH}$ and applying this to
proposed values of set of underlying pKs using data obtained previously
[@Ring2023-acid]. Next, it will be examined if data generated with
specified $\mathrm{p}K$ values over extended or restricted range of
$\mathrm{pH}$ values allows recovery of these same $\mathrm{p}K$ values.
Accepting that $\mathrm{pH} = -1$ would hardly be realized in biological
fluid and, anyway would require dramatic ionic strengths invalidating
conventional understanding of $\mathrm{pH}$ modeling, the consequence
for modeling of assuming the first $\mathrm{p}K_1$ fixed at different
low values will be examined. In so far as the developed model for
$\mathrm{pH}$ dependent charge on HEPES implies a strict ordering of
$\mathrm{p}K$ values, as will be seen in the Methods section, the
consequence for recovery of known $\mathrm{p}K$ values of assuming
strict ordering will be assessed. The final model, thus developed, was
used with the original observed data to obtain best values for the
underlying $\mathrm{p}K$ values and the fit assessed. Finally, the
sensitivity of the final model to imposed noise in $\mathrm{SID}$,
$\mathrm{[HEPES]}$, or $\mathrm{pH}$ will be presented.

Methods
=======

Assuming that $\mathrm{p}K_{a1}$ of the sulfonic group is much lower
than those on the piperazine nitrogens as also supported by conversation
on StackExchange [@ChemSE:HEPES-pK], we will have the following sequence
moving from low to high $\mathrm{pH}$:

```{r echo=FALSE}
# A dash! A dash! My kingdom for a dash!
# (The things that rmarkdown users have to do without
# \usepackage{mhchem})
```
$\mathrm{I: H^{+}N_1\text-HEP\text-N_4H^{+}SO_3H}$ $\sim \mathrm{p}K_{a1}$ on
sulfonic group

$\mathrm{II: H^{+}N_1\text-HEP\text-N_4H^{+}SO_3^{-}}$ $\sim \mathrm{p}K_{a2}$ on
$\mathrm{N_1}$

$\mathrm{III: N_1\text-HEP\text-N_4H^{+}SO_3^{-}}$ $\sim \mathrm{p}K_{a3}$ on
$\mathrm{N_4}$

$\mathrm{IV: N_1\text-HEP\text-N_4SO_3^{-}}$

Now $\mathrm{I + II + III + IV = HEPTOT}$

From the individual dissociations we have

\[
\mathrm{IV} = \frac{
 \mathrm{III} \cdot 10^{-\mathrm{p}K_{a3}}
}{
 [\mathrm{H}^{+}]
} = \frac{
 \mathrm{II} \cdot 10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}}
}{
 [\mathrm{H}^{+}]^2
} = \frac{
 \mathrm{I}
  \cdot 10^{-\mathrm{p}K_{a3}}
  \cdot 10^{-\mathrm{p}K_{a2}}
  \cdot 10^{-\mathrm{p}K_{a1}}
}{
 [\mathrm{H}^{+}]^3
}
\]

\[
\mathrm{I} = \frac{\mathrm{IV} [\mathrm{H}^{+}]^3}{
 10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}} \cdot 10^{-\mathrm{p}K_{a1}}
}
\]

\[
\mathrm{II} = \frac{\mathrm{IV} [\mathrm{H}^{+}]^2}{
 10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}}
}
\]

\[
\mathrm{III} = \frac{\mathrm{IV} [\mathrm{H}^{+}]}{
 10^{-\mathrm{p}K_{a3}}
}
\]

\[
\mathrm{I} + \mathrm{II} + \mathrm{III} = \mathrm{IV} \cdot \left[
 \frac{[\mathrm{H}^{+}]^3}{
  10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}} \cdot 10^{-\mathrm{p}K_{a1}}
 }
 +
 \frac{[\mathrm{H}^{+}]^2}{
  10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}}
 }
 +
 \frac{[\mathrm{H}^{+}]}{
  10^{-\mathrm{p}K_{a3}}
 }
\right] = \mathrm{HEPTOT} - \mathrm{IV}
\]

\[
\mathrm{IV} = \frac{\mathrm{HEPTOT}}{
 1 + \left[
  \frac{[\mathrm{H}^{+}]^3}{
   10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}} \cdot 10^{-\mathrm{p}K_{a1}}
  }
  +
  \frac{[\mathrm{H}^{+}]^2}{
   10^{-\mathrm{p}K_{a3}} \cdot 10^{-\mathrm{p}K_{a2}}
  }
  +
  \frac{[\mathrm{H}^{+}]}{
   10^{-\mathrm{p}K_{a3}}
  }
 \right]
}
\]

So, we can see that $\mathrm{IV}$ is identified from total HEPES
concentration at a given $\mathrm{pH}$ if the three $\mathrm{p}K$ values
are known, and the derivation clearly requires that $\mathrm{p}K_{a1} <
\mathrm{p}K_{a2} < \mathrm{p}K_{a3}$. Consequently, we have quantified
the concentration of all HEPES moieties. Furthermore, the net charge on
each of the four HEPES components are known: $\mathrm{I}: +2$;
$\mathrm{II}: +1$; $\mathrm{III}: 0$; $\mathrm{IV}: -1$. Thereby we have
obtained a specification for the $\mathrm{pH}$ dependent net charge on
HEPES given its total concentration and given the three named
$\mathrm{p}K$ values.  Therefore, with SID denoting strong ion
difference[@Ring2022] and $K_w$ the auto-dissociation of water, the new
charge-balance model for HEPES is:

\begin{equation}
[\mathrm{H}^{+}] + SID - \frac{K_w}{[\mathrm{H}^{+}]} + \mathrm{I} \cdot 2
+ \mathrm{II} - \mathrm{IV} = 0
\end{equation}

With the three $\mathrm{p}K_a$ values for HEPES taken to be $-1$, $3$
and $7.55$ this will be used to estimate $\mathrm{pH}$ in experimental
fluids to compare with observed $\mathrm{pH}$. It should be noted that
Kildeberg[@Kildeberg1983] derives for any multiprotic substance the
average charge given an ordered sequence of $\mathrm{p}K_1 <
\mathrm{p}K_2 \dots < \mathrm{p}K_n$ as
$N = \frac{
 10^{\mathrm{pH} - \mathrm{p}K_1} +
 2 \cdot 10^{2 \mathrm{pH} - \mathrm{p}K_1 - \mathrm{p}K_2} +
 \dots +
 k \cdot  10^{k \mathrm{pH} - \mathrm{p}K_1 \dots \mathrm{p}K_k}
}{
 1+
 10^{\mathrm{pH} - \mathrm{p}K_1} +
 \cdot 10^{2 \mathrm{pH} - \mathrm{p}K_1 - \mathrm{p}K_2} +
 \dots +
 \cdot  10^{k \mathrm{pH} - \mathrm{p}K_1 \dots \mathrm{p}K_k}
} - n$
with $n$ denoting the charge on the fully protonated species. For given
$\mathrm{p}K$ values and $n$ this yields exactly the same charge as
derived above, but since we are also interested in inverse inference
about the $\mathrm{p}K$ values based on observed chemistry and
$\mathrm{pH}$, the first formulation derived here is more malleable when
specifying required order of $\mathrm{p}K$ entries.

Next, using the observed $\mathrm{pH}$ it was attempted to recover the
$\mathrm{p}K_a$ values for HEPES based on known chemical composition.
This included generating _in silico_ data with known $\mathrm{p}K_a$
values for HEPES and recovering again the $\mathrm{p}K_a$ values from
the synthetic data and examining the sensitivity of parameter
estimates/recovery to changing $\mathrm{pH}$ by modification of SID or
by changing HEPES concentration. The $\mathrm{p}K_a$ values for HEPES
were found from non-linear regression using **R** package `nlsr`
[@CRAN:nlsr] and the ordering of the $\mathrm{p}K_a$ values were
enforced by a difference method documented in supplementary materials.

References
==========

### 2023-11-15 manuscript.Rmd
Added reference to Nash/Bhattacharjee and an Appendix about methods.

@@
 
+Appendix: Software considerations
+=================================
+
+The `nlsr` package @CRAN:nlsr which has been used for the estimation of
+parameters in the nonlinear models for this article has some features that 
+more easily permitted the computations. Moreover, discussions concerning how
+to use the software led to unexpected possibilities in that use.
+
+### Ensuring parameters follow a specific ordering
+
+We require that the three $\mathrm{p}K_a$ values be in increasing order of
+numerical value. To simplify the discussion here, we will write these as 
+`p1`, `p2` and `p3` within our R program scripts. Thus we want
+
+```
+       p1  <  p2  < p3
+```
+
+One approach, which we have used here, is to reparameterize the formulas for
+the model so
+
+``` 
+      p2  =  p1 + d12
+      p3  =  p1 + d12 + d23
+```
+
+If we constrain `d12` and `d23` to be positive, or at least non-negative, then
+the new parameterization will ensure the ordering. However, this does require 
+that the software can offer bounds constraints on the parameters. `nslr` is 
+capable in this way, and even allows upper and lower bounds to be equal, 
+resulting in fixed parameters, or **masks**. Other software in R has similar
+capabilities, but the built-in `nls()` function only handles bounds if the
+`port` algorithm is used, and there are warnings that this code may be 
+incomplete. Package `minpack.lm` also handles bounds, though a failure was
+encountered in one example (@NashBhattacharjee).
+
+### Vectorizing a function ???
+
+JN: As yet I don't understand the workings of this, and must admit to never
+having used Vectorize(). I think it worthwhile giving a good explanation.
+
+### Gauging the reliability of the estimated parameters
+
+All parameter estimation methods must contend with the possibility that 
+a variety of parameter values may give very similar magnitudes for the loss
+function (e.g, sum of squared residuals) that is optimized to obtain the
+estimates. In the case of estimation that can be converted to minimizing a
+sum of squares, the Jacobian of the vector of residuals will become close 
+to singular if parameters are indeterminate. This can be judged by computing
+the **singular values** (??JN: do we want a reference) of the Jacobian and
+package `nlsr` does this and displays the values when a summary of a solution 
+is requested. A rule of thumb is to regard the Jacobian as essentially 
+singular if the ratio of the smallest to largest singular value is "small".
+A value of smaller than 1e-4 would be a typical criterion. Note that more than
+one singular value may be small, indicating that there is more than one 
+dimension of indeterminacy. However, such indicators apply to the ensemble of
+parameters, and as such concern the overall model.
+
 References

### 2023-11-17 manuscript.Rmd
Vectorize: partially describe the problem

@@
 
-### Vectorizing a function ???
+### Vectorizing a function
 
-JN: As yet I don't understand the workings of this, and must admit to never
-having used Vectorize(). I think it worthwhile giving a good explanation.
+**IK**: This feels like I'm writing a tutorial. Should I be less
+detailed?
+
+**R** scripts naturally lend themselves to vectorization, thanks to the
+language performing many operations on whole vectors. For example, **R**
+has no problem evaluating the charge-balance equation function for a
+vector of different values of $[\mathrm H]$ with other values remaining
+scalars:
+
+```{r HEPESFUNC}
+HEPESFUNC <- function(H,SID,HEPTOT,pK1,pK2,pK3) {
+	XX <- (H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
+	IV <- HEPTOT/(1+XX)
+	I <- IV*H^3/(10^-pK3*10^-pK2*10^-pK1)
+	II <- IV*H^2/(10^-pK3*10^-pK2)
+	III <- IV*H/10^-pK3
+	H - kw/H + SID + I*2 + II - IV
+}
+```
+
+This principle is used by the non-linear regression functions which, when
+given a formula `y ~ f(x, theta)` and vectors `x` of independent
+variables and `y` of observed values of the dependent variable, call the
+model function `f` with the whole vector `x` as argument, expecting to
+receive an estimate of the whole `y` at the current value of the
+parameter `theta` in response.
+
+Some functions are not naturally vectorized and cannot be easily changed
+to obtain this property. For example, the model function for a
+non-linear regression problem may involve root-finding using `uniroot`:
+
+```{r}
+# Which pH value satisfies the charge-balance equation?
+pH <- function(SID, HEPTOT, pK1, pK2, pK3)
+  -log10(uniroot(HEPESFUNC,c(1e-20,1),tol=1e-20,maxiter=1e4,
+                 HEPTOT=HEPTOT,SID = SID, pK1=pK1,pK2=pK2,pK3=pK3)$root)
+```
+
+The underlying mechanism of `uniroot()` only works with scalar functions
+of one argument, so passing vectors of length more than `1` to this
+function will cause an error: `uniroot()` is not prepared to find
+multiple different roots for every given set of the parameter values. In
+order to use this function with non-linear regression methods, a wrapper
+is needed. This wrapper must call the original function in a loop for
+every element of the vector arguments passed to it:
+
+```{r}
+pH.wrapper <- function(SID, HEPTOT, pK1, pK2, pK3)
+ vapply(SID, function(SID_i)
+  -log10(uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
+                 HEPTOT = HEPTOT, SID = SID_i,
+                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root),
+  numeric(1)
+ )
+```
+
+The base function `Vectorize` automates this transformation. It returns
+a wrapper function that obtains the values of every argument given to
+the wrapper and passes them to a live-constructed call to `mapply()`,
+letting the latter call the original function. A call to
+`mapply(f, a, b, c)`, returns a vector of values of
+`f(a[i], b[i], c[i])` for every applicable `i`, subject to vector
+recycling rules.
 

### 2023-11-17 manuscript.Rmd
Illustrate Vectorize()

@@
 
-**IK**: This feels like I'm writing a tutorial. Should I be less
+**IK**: This feels like I'm writing a tutorial. Should it be less
 detailed?
@@
 ```{r HEPESFUNC}
+kw <- 1e-14
 HEPESFUNC <- function(H,SID,HEPTOT,pK1,pK2,pK3) {
@@
 `f(a[i], b[i], c[i])` for every applicable `i`, subject to vector
-recycling rules.
+recycling rules:
+
+```{r}
+pHm <- Vectorize(pH)
+ans <- c(
+ pH(.04, .05, -1, 3, 7.55),
+ pH(.035, .05, -1, 3, 7.55)
+)
+all.equal(ans, pH.wrapper(
+ c(.04, .035), .05, -1, 3, 7.55
+))
+all.equal(ans, pHm(
+ c(.04, .035), .05, -1, 3, 7.55
+))
+```
 

### 2023-11-17 manuscript.Rmd
Vectorize: differentiating past root-finding
It's actually possible, but takes certain gymnastics in both R and
calculus, to obtain the symbolic Jacobian despite a root-finding
operation in the middle.

@@
 
+Unfortunately, the use of automatic vectorization precludes the use of
+fully automatic symbolic differentiation: static code analysis will not
+go past `vapply`, `mapply`, or `uniroot`. Still, it is possible to
+construct most of the expression piece by piece and have it
+differentiated this way:
+
+```{r}
+charge_balance <- local({
+	XX <- quote(H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
+	IV <- substitute(HEPTOT/(1+XX))
+	I <- substitute(IV*H^3/(10^-pK3*10^-pK2*10^-pK1))
+	II <- substitute(IV*H^2/(10^-pK3*10^-pK2))
+	III <- substitute(IV*H/10^-pK3)
+	nlsr::codeDeriv(
+		substitute(H - kw/H + SID + I*2 + II - IV),
+		c('H', 'pK1', 'pK2', 'pK3')
+	)
+})
+# a few screenfuls of code; >2500 pages of output with verbose=TRUE
+```
+
+In order to obtain the final Jacobian, we will have to somehow
+differentiate past the root-finding operation. For that we will use the
+implicit function theorem. Let the charge-balance equation be
+$f([\mathrm H], \theta) = 0$, with $[\mathrm H]$ being the desired
+equilibrium concentration of protons and $\theta$ symbolising all of the
+parameters of the equation. It implicitly defines the function
+$[\mathrm H]_\mathrm{eq}(\theta)$ that satisfies
+$F(\theta) = f([\mathrm H]_\mathrm{eq}(\theta), \theta) = 0$. Differentiating the
+latter equation by $\theta$, we get $\frac{dF}{d\theta} =
+\frac{\partial{}f}{\partial[\mathrm H]}
+\frac{d[\mathrm H]_\mathrm{eq}}{d\theta} +
+\frac{\partial{}f}{\partial\theta} \frac{d\theta}{d\theta} = 0$, from
+which we get $\frac{d[\mathrm H]_\mathrm{eq}}{d\theta} =
+-\frac{\partial{}f/\partial\theta}{\partial{}f/\partial[\mathrm H]}$:
+
+```{r}
+pH.sym <- function(SID, HEPTOT, pK1, pK2, pK3) {
+ H <- vapply(SID, function(SID) { # for every point...
+  # find the root of the function
+  H <- uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
+                 HEPTOT = HEPTOT, SID = SID,
+                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root
+  # evaluate the derivatives by H and parameters
+  g <- attr(eval(charge_balance), 'gradient')
+  # and calculate the derivative of the H_eq by parameters;
+  # combine together for easier access by vapply
+  c(H, -g[,2:4]/g[,1])
+ }, numeric(4)
+ )
+ # now take the negative logarithm to obtain pH
+ structure(-log10(H[1,]), gradient = -(1/(H[1,] * log(10))) * t(H[-1,]))
+}
+pH.sym(
+ c(.04, .035), .05, -1, 3, 7.55
+)
+```
+
+This gets similar, but not exactly the same results as numeric
+differentiation:
+
+```{r}
+nlsr::numericDerivR(
+ quote(pHm(
+  c(.04, .035), .05, pK1, pK2, pK3
+ )),
+ c('pK1', 'pK2', 'pK3'),
+ list2env(list(pK1 = -1, pK2 = 3, pK3 = 7.55)),
+ central = TRUE
+)
+```
+
 ### Gauging the reliability of the estimated parameters

### 2023-11-18 manuscript.Rmd
Formatting tweaks

@@
 \begin{equation}
-[\mathrm{H}^{+}] + SID - \frac{K_w}{[\mathrm{H}^{+}]} + \mathrm{I} \cdot 2
-+ \mathrm{II} - \mathrm{IV} = 0
+[\mathrm{H}^{+}] + \mathrm{SID} - \frac{K_w}{[\mathrm{H}^{+}]}
++ \mathrm{I} \cdot 2 + \mathrm{II} - \mathrm{IV} = 0
 \end{equation}
@@
 ```{r}
-pHm <- Vectorize(pH)
 ans <- c(
@@
 )
-all.equal(ans, pH.wrapper(
- c(.04, .035), .05, -1, 3, 7.55
-))
-all.equal(ans, pHm(
- c(.04, .035), .05, -1, 3, 7.55
-))
+all.equal(ans, pH.wrapper(c(.04, .035), .05, -1, 3, 7.55))
+pHm <- Vectorize(pH)
+all.equal(ans, pHm(c(.04, .035), .05, -1, 3, 7.55))
 ```
@@
 pH.sym <- function(SID, HEPTOT, pK1, pK2, pK3) {
- H <- vapply(SID, function(SID) { # for every point...
+ H <- vapply(SID, function(SID) { # for every point:
   # find the root of the function
@@
   c(H, -g[,2:4]/g[,1])
- }, numeric(4)
- )
+ }, numeric(4))
  # now take the negative logarithm to obtain pH
@@
 }
-pH.sym(
- c(.04, .035), .05, -1, 3, 7.55
-)
+pH.sym(c(.04, .035), .05, -1, 3, 7.55)
 ```
 
-This gets similar, but not exactly the same results as numeric
+This gives similar, but not exactly the same results as numeric
 differentiation:
@@
 nlsr::numericDerivR(
- quote(pHm(
-  c(.04, .035), .05, pK1, pK2, pK3
- )),
+ quote(pHm(c(.04, .035), .05, pK1, pK2, pK3)),
  c('pK1', 'pK2', 'pK3'),

### 2023-11-18 manuscript.Rmd
Least-squares solution of charge-balance equations
Instead of solving the charge-balance equation at every known SID, we
can use the least squares method to solve them all together, minimising
the sum of squared change imbalances.

This gives us worse results due to a different error structure of the
model, but may be worth noting.

@@
 + \mathrm{I} \cdot 2 + \mathrm{II} - \mathrm{IV} = 0
+(\#eq:charge-balance)
 \end{equation}
@@
 pH.wrapper <- function(SID, HEPTOT, pK1, pK2, pK3)
- vapply(SID, function(SID_i)
+ mapply(function(SID, HEPTOT)
   -log10(uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
-                 HEPTOT = HEPTOT, SID = SID_i,
+                 HEPTOT = HEPTOT, SID = SID,
                  pK1 = pK1, pK2 = pK2, pK3 = pK3)$root),
-  numeric(1)
+  SID = SID, HEPTOT = HEPTOT
  )
@@
 
+### Solving the charge-balance equations in the least squares sense
+
+Yet another way to model the data is to plug the experimental values of
+$\mathrm{SID}$ and measured $\mathrm{pH}$ straight into the
+charge-balance equation \@ref(eq:charge-balance), without going through
+the intermediate step of finding the negative logaritm of the root of
+the function. This changes the problem: instead of
+$\min_{\mathrm{p}K_{a,1}, \mathrm{p}K_{a,2}, \mathrm{p}K_{a,3}}
+\sum_i(\mathrm p\hat{\mathrm H}(\mathrm{SID}_i, \mathrm{p}K_{a,1},
+\mathrm{p}K_{a,2}, \mathrm{p}K_{a,3}) - \mathrm{pH}_i)^2$, we are now
+minimising the sum of squared residuals from eq.
+\@ref(eq:charge-balance) itself:
+
+```{r}
+load('HEPES.RData') # provides the DD data.frame
+charge_balance_fn <- function() NULL
+formals(charge_balance_fn) <- formals(HEPESFUNC)
+body(charge_balance_fn) <- local({
+	XX <- quote(H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
+	IV <- substitute(HEPTOT/(1+XX))
+	I <- substitute(IV*H^3/(10^-pK3*10^-pK2*10^-pK1))
+	II <- substitute(IV*H^2/(10^-pK3*10^-pK2))
+	III <- substitute(IV*H/10^-pK3)
+	nlsr::codeDeriv(
+		substitute(H - kw/H + SID + I*2 + II - IV),
+		c('pK1', 'pK2', 'pK3')
+	)
+})
+
+loss <- function(x) charge_balance_fn(
+	H = 10^-DD$OBS, SID = DD$SID, HEPTOT = DD$HEPES,
+	pK1 = x['pK1'], pK2 = x['pK2'], pK3 = x['pK3']
+)
+(m_cb <- nlsr::nlfb(
+	start = c(pK1 = -1, pK2 = 3, pK3 = 7.55),
+	resfn = loss, jacfn = loss,
+	lower = -3, upper = 9
+))
+```
+
+While this avoids finding the roots of the charge-balance equation, the
+implied residual structure (normally-distributed errors in the
+$\mathrm{pH}$ measurements in the original model vs.
+normally-distributed charge imbalances in concentration units in the
+current model) is different, and so are the resulting estimates:
+
+```{r}
+(m_rf <- nlsr::nlxb(
+	OBS ~ pH.wrapper(SID, HEPES, pK1, pK2, pK3),
+	data = DD, start = c(pK1 = -1, pK2 = 3, pK3 = 7.55),
+	lower = -3, upper = 9,
+	control = list(japprox = 'jacentral')
+))
+```
+
 ### Gauging the reliability of the estimated parameters

### 2023-11-28 noise.R
Perform the noise/vector length experiment
The code now has two more direct dependencies (pbapply & depcache) but
can be parallelised and restarted after an interruption.

The experiment was also made smaller by dropping some levels.

@@ -1,3 +1,8 @@
+library(parallel)
+library(pbapply)
+loadNamespace('depcache')
+stopifnot(`Please run this script with a cluster object registered` = !is.null(cl <- getDefaultCluster()))
+
 source("definitions.R")
 kw <- 1e-14
 HEPTOT <- 0.05
@@ -7,33 +12,46 @@ RNGversion('3.6.0')
 set.seed(1365)
 
 params <- expand.grid(
-	pK_true = replicate(16, sort(runif(3, -3, 9)), FALSE),
-	replication = 1:16,
+	replication = seq_len(10),
+	pK_true = replicate(10, sort(runif(3, -3, 9)), FALSE),
 	noise_ratio = 10^seq(-4, -.5, len = 8),
-	n_points = c(10, 20, 50, 100, 200, 500, 1000, 2000, 5000, 10000),
 	noise = list(
 		normal = rnorm,
 		Cauchy = rcauchy,
 		uniform = function(n) runif(n, -1, 1)
-	)
+	),
+	n_points = rev(c(10, 20, 50, 100, 200, 500, 1000, 2000))
 )
 
+getrows <- function(d) lapply(seq_len(nrow(d)), function(i) d[i,])
 vnorm <- function(x) sqrt(sum(x^2))
 
-params$m <- lapply(seq_len(nrow(params)), function(i) {
-	p <- params[i,]
+# Set the noise in stone before we start using parallel operations.
+# Far from the most efficient way to do this, but at least it's reproducible.
+params$vnoise <- pblapply(getrows(params), function(p) p$noise[[1]](p$n_points))
+# Send over all variables except the giant 'params' and the cluster itself.
+clusterExport(cl, setdiff(ls(), c('cl','params')))
+#pboptions(use_lb = TRUE)
+
+params$m <- pblapply(getrows(params), function(p) {
 	SID <- seq(-5e-2, 4e-2, length.out = p$n_points)
 	pH <- pHm(SID, HEPTOT, p$pK_true[[1]][1], p$pK_true[[1]][2], p$pK_true[[1]][3])
-	noise <- p$noise[[1]](length(pH))
-	pH <- pH + noise * vnorm(pH) / vnorm(noise) * p$noise_ratio
-	nlsr::nlxb(
+	pH <- pH + p$vnoise[[1]] * vnorm(pH) / vnorm(p$vnoise[[1]]) * p$noise_ratio
[…]

### 2023-11-30 noise.Rmd
Draw the plots from the noise experiment

First let's load the results produced by `noise.R` overnight.

The results are organised a bit clumsily, with `pK_true` being a list of
numeric vectors. Our goal is to group by every experiment condition (set
of true $\mathrm{p}K_a$ values, number of points, noise-to-signal ratio,
kind of noise) and summarise the $10$ repeats that were conducted with
different samples of noise.

```{r}
n <- readRDS('noise.rds')
# On its own, `by` cannot group on list columns, so produce a surrogate
# key for the true pK column, representing numeric vectors as strings.
# This is probably not efficient at all, but will work for our purposes.
n$idx <- as.factor(sapply(n$pK_true, paste, collapse = ' '))
levels(n$idx) <- seq_along(levels(n$idx))
pn <- do.call(rbind, by(
	n, n[c('noise_ratio','n_points','noise_kind', 'idx')],
	function(d) {
		pK_true <- d$pK_true[[1]]
		# combine all estimated pKs into one matrix
		pK_est <- simplify2array(d$pK_est)
		d$pK_est <- NULL
		cbind(d[1,],
			pK_est = I(list(t(pK_est))),
			# average residual: should approach 0 for unbiased estimator
			res_mean = mean(pK_est - pK_true),
			# deviance of residuals: measures the spread of the
			# estimates
			res_dev = mean((pK_est - pK_true)^2)
		)
	}
))
```

Every row of the newly obtained `data.frame` now contains results for
every repeat of the same set of experimental conditions:

```{r}
head(pn)
pn[[1,'pK_est']]
```

Let's reduce it further, averaging every generated set of true
$\mathrm{p}K_a$ values:

```{r}
pna <- aggregate(
	pn[c('res_mean','res_dev')],
	pn[c('noise_ratio','n_points','noise_kind')],
	mean
)
```

This lets us plot the average estimation error and the standard
deviation of the estimates:
```{r}
lattice::levelplot(
	res_mean ~ noise_ratio + n_points | noise_kind, pna,
	group = noise_kind, scales = list(log = 10),
	main = quote(
		'Average '*pK[a]*' estimation error '*
		'(should be close to 0)'
	),
	xlab = 'Noise-to-signal ratio',
	ylab = 'Number of known points',
	cuts = 255, col.regions = hcl.colors(256, 'Broc')
)
```

I don't see any pattern in the data, and it seems to be fairly centered
around $0$, so the estimate is effectively unbiased, even in presence of
non-normally-distributed noise. But what about the variance?
```{r}
lattice::levelplot(
	sqrt(res_dev) ~ noise_ratio + n_points | noise_kind, pna,
	group = noise_kind, scales = list(log = 10),
	main = quote(
		'Standard deviation of '*pK[a]*' estimate'
	),
	xlab = 'Noise-to-signal ratio',
	ylab = 'Number of known points',
	cuts = 255, col.regions = hcl.colors(256, 'Terrain 2')
)
```

Given $10^{1.5}$ points and realistic (?) noise, the standard deviation
of the $\mathrm{p}K_a$ estimates can be as high as $1.5$. Under
three-sigma rule, this gives us an error of up to $4.5$ $\mathrm{pH}$
units...

### 2023-12-01 Rosevec.R
update Rosevec

@@ -1,0 +1,22 @@
+# Rosevec.R -- try to vectorize Rosenbrock function
+# 20231201 -- Thanks to Ivan for sorting this out for me, JN
+onerose<-function(i,x, scale){
+   if ( ! (i %in% 1:2)) stop("bad index i")
+   if (i == 1) {
+       res <- scale*(x[2] - x[1]^2)
+   } else {
+       res <- (1 - x[1])
+   }
+   res
+}
+
+onerose_v <- Vectorize(onerose, 'i')
+allres <- function(x, scale) onerose_v(1:2, x, scale)
+x <- c(-1.2, 1)
+library(adagio)
+print(x)
+cat("adagio::fnRosenbrock(x):",fnRosenbrock(x),"\n")
+aa <- allres(x, scale=10)
+print(aa)
+print(sum(aa^2))
+

### 2023-12-01 noise.Rmd

@@
 Every row of the newly obtained `data.frame` now contains results for
-every repeat of the same set of experimental conditions:
+all repeats of the same set of experimental conditions:
 
@@
 ```{r}
-lattice::levelplot(
+library(lattice)
+levelplot(
 	res_mean ~ noise_ratio + n_points | noise_kind, pna,
@@
 ```{r}
-lattice::levelplot(
+levelplot(
 	sqrt(res_dev) ~ noise_ratio + n_points | noise_kind, pna,
@@
 	ylab = 'Number of known points',
-	cuts = 255, col.regions = hcl.colors(256, 'Terrain 2')
+	cuts = 255, col.regions = hcl.colors(256, 'Terrain 2'),
+	panel = function(x, y, z, subscripts, at, contour, region, ...) {
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts,
+			at = at, contour = FALSE, region = TRUE, ...
+		)
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts, region = FALSE,
+			at = c(1,2,3), contour = TRUE,
+			labels = list(cex = .75), label.style = 'align', ...
+		)
+	}
 )
@@
 
-Given $10^{1.5}$ points and realistic (?) noise, the standard deviation
-of the $\mathrm{p}K_a$ estimates can be as high as $1.5$. Under
-three-sigma rule, this gives us an error of up to $4.5$ $\mathrm{pH}$
-units...
+What about the individual $\mathrm{p}K_a$ values?
+
+```{r, fig.height = 10}
+pnw <- cbind(pna, t(mapply(function(est, true) {
+	rowMeans((t(est) - true)^2)
+}, pn$pK_est, pn$pK_true)))
+levelplot(
+	sqrt(pK1) + sqrt(pK2) + sqrt(pK3) ~ noise_ratio + n_points | noise_kind,
+	aggregate(
+		pnw[c('pK1','pK2','pK3')],
+		pnw[c('noise_ratio','n_points','noise_kind')],
+		mean
+	),
+	group = noise_kind, scales = list(log = 10),
+	main = quote(
+		'Standard deviation of '*pK[a]*' estimate'
+	),
+	xlab = 'Noise-to-signal ratio',
+	ylab = 'Number of known points',
+	cuts = 255, col.regions = hcl.colors(256, 'Terrain 2'),
+	panel = function(x, y, z, subscripts, at, contour, region, ...) {
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts,
+			at = at, contour = FALSE, region = TRUE, ...
+		)
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts, region = FALSE,
+			at = pretty(z), contour = TRUE,
+			labels = list(cex = .75), label.style = 'align', ...
+		)
+	}
+)
+```
+
+One more way to quantify the error. Let's take a look at every model
+fitted so far, group by experiment conditions (noise-to-signal ratio,
+number of points, noise distribution) and obtain the 95% highest
+absolute deviation between estimated and true value of individual
+$\mathrm{p}K_a$s. This will save us from having to recalculate standard
+deviations of the estimates into confidence bands. Bonus points if the
+error turs out to be independent from the noise distribution.
+
+```{r, fig.height = 10}
+nw <- cbind(n, t(abs(mapply(`-`, n$pK_est, n$pK_true))))
+levelplot(
+	pK1 + pK2 + pK3 ~ noise_ratio + n_points | noise_kind,
+	aggregate(
+		nw[c('pK1','pK2','pK3')],
+		nw[c('noise_ratio','n_points','noise_kind')],
+		quantile, .95
+	),
+	group = noise_kind, scales = list(log = 10),
+	main = quote(
+		{95^'th'}*' percentile of absolute '*pK[a]*' estimation error'
+	),
+	xlab = 'Noise-to-signal ratio',
+	ylab = 'Number of known points',
+	cuts = 255, col.regions = hcl.colors(256, 'Terrain 2'),
+	panel = function(x, y, z, subscripts, at, contour, region, ...) {
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts,
+			at = at, contour = FALSE, region = TRUE, ...
+		)
+		panel.levelplot(
+			x = x, y = y, z = z, subscripts = subscripts, region = FALSE,
+			at = pretty(z), contour = TRUE,
+			labels = list(cex = .75), label.style = 'align', ...
+		)
+	}
+)
+```
+
+Hard to say whether the noise distribution really makes a difference,
+but $\mathrm{p}K_{a3}$ can be estimated reliably, $\mathrm{p}K_{a2}$ may
+require a lot of points and good agreement between experiment and model,
+and $\mathrm{p}K_{a1}$ is absolutely hopeless. But we knew about the latter.

### 2023-12-08 noise.Rmd
See if estimation error depends on pKa/pH vicinity

@@
 and $\mathrm{p}K_{a1}$ is absolutely hopeless. But we knew about the latter.
+
+Does the estimation error depend on the vicinity between the observed
+$\mathrm{pH}$ and the true value of $\mathrm{p}K_a$?
+
+```{r, fig.height = 8}
+source('definitions.R')
+kw <- 1e-14
+HEPTOT <- 0.05
+nw <- cbind(nw, vicinity = t(mapply(
+	function(pK_true, n_points) {
+		SID <- seq(-5e-2, 4e-2, n_points)
+		pK_true - mean(pHm(
+			SID, HEPTOT, pK_true[1], pK_true[2], pK_true[3]
+		))
+	}, nw$pK_true, nw$n_points
+)))
+p1 <- xyplot(
+	sqrt(pK1) ~ vicinity.1,
+	aggregate(
+		cbind(nw[c('pK1')]^2, nw['vicinity.1']),
+		nw[c('noise_ratio','n_points','noise_kind','idx')],
+		mean
+	),
+	subset = n_points <= 100 & noise_ratio <= 1e-2,
+	group = paste(noise_ratio, n_points, noise_kind),
+	scales = list(y = list(log = 10)),
+	xlab = quote(pK[1]^'true' - mean(pH)),
+	ylab = quote(sigma(pK[1]))
+)
+p2 <- xyplot(
+	sqrt(pK2) ~ vicinity.2,
+	aggregate(
+		cbind(nw[c('pK2')]^2, nw['vicinity.2']),
+		nw[c('noise_ratio','n_points','noise_kind','idx')],
+		mean
+	),
+	subset = n_points <= 100 & noise_ratio <= 1e-2,
+	group = paste(noise_ratio, n_points, noise_kind),
+	scales = list(y = list(log = 10)),
+	xlab = quote(pK[2]^'true' - mean(pH)),
+	ylab = quote(sigma(pK[2]))
+)
+p3 <- xyplot(
+	sqrt(pK3) ~ vicinity.3,
+	aggregate(
+		cbind(nw[c('pK3')]^2, nw['vicinity.3']),
+		nw[c('noise_ratio','n_points','noise_kind','idx')],
+		mean
+	),
+	subset = n_points <= 100 & noise_ratio <= 1e-2,
+	group = paste(noise_ratio, n_points, noise_kind),
+	scales = list(y = list(log = 10)),
+	xlab = quote(pK[3]^'true' - mean(pH)),
+	ylab = quote(sigma(pK[3]))
+)
+print(p1, split = c(1, 1, 1, 3), more = TRUE)
+print(p2, split = c(1, 2, 1, 3), more = TRUE)
+print(p3, split = c(1, 3, 1, 3))
+```
+
+It's very hard to see a pattern, and we have too much variability in
+kinds of noise and not enough variability in true $\mathrm{p}K_a$
+values, but it seems that if the true value of $\mathrm{p}K_a$ exceeds
+the $\mathrm{pH}$ value on average, it can be reliably estimated.

### 2024-02-04 manuscript.Rmd
Import edits up to figure 2

@@
 
-```{r, echo=FALSE}
-# NOTE: bookdown features (e.g. cross-references) used below
-# This requires the use of a bookdown output format, e.g.
-# bookdown::pdf_document2 or bookdown::html_document2
-```
+<!--
+NOTE: bookdown features (e.g. cross-references) used below
+This requires the use of a bookdown output format, e.g.
+bookdown::pdf_document2 or bookdown::html_document2
+-->
 
@@
 
-It has been shown [@Ring2022] that $\mathrm{pH}$ in arbitrary fluids can be
-understood based on the three fundamental principles of
-electroneutrality, conservation of mass, and rules of dissociation as
-devised from physical chemistry. It has also been shown [@Ring2022] that the
-charge-balance approach allows definition of the $\mathrm{pH}$-dependent
-buffer-capacity, and provides a definition of acid content as titratable
-acidity, based on previous work by Kildeberg [@Kildeberg1983]. Recently
-it has been reiterated [@Ring2023-charge] that it is absolutely necessary for this
-modeling approach to take into account the strong ions, which are those
-assumed to be completely dissociated at any $\mathrm{pH}$. In contrast to the
-physico-chemical model of Stewart [@Stewart1978], the charge-balance
-model (CBM, henceforth) has no separation between independent and
-dependent variables. Under the CBM all that can be known about the
-system is gathered in the expression of charge in an explicitly stated
-chemical system. Consequently, if, as an example, $\mathrm{pH}$ is known and charge
-is a known deterministic function of the chemical specification, inverse
-inferences on unknown variables can, at least sometimes, be obtained by
-root finding, since we know that the charge must be zero. In simple
-systems, such as a single monovalent weak acid and strong ions, a single
-unknown variable is explicitly estimable [@Ring2022]. The question
-focused here is if CBM is rich enough to allow inverse inference in more
-complex systems such as with multiprotic buffers. This question arose
-under experiments with HEPES buffered sulfate containing fluids
-[@Ring2023-acid] in which HEPES was modeled as a monovalent buffer with
-pKa 7.5 [@Ring2010] since it was found that estimated $\mathrm{pH}$ was much too
-low when observed $\mathrm{pH}$ was less than about 6. HEPES is a zwitterion with
-formula 2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic acid and
-structure as shown in \@ref(fig:hepes).
+It has been shown[@Ring2022] that $\mathrm{pH}$ in arbitrary fluids can be
+understood based on the three fundamental principles of electroneutrality,
+conservation of mass, and rules of dissociation as devised from physical
+chemistry. It has also been shown[@Ring2022] that the charge-balance approach
+allows definition of the pH-dependent buffer-capacity, and provides a definition
+of acid content as titratable acidity, based on previous work by
+Kildeberg[@Kildeberg1983]. Recently it has been reiterated[@Ring2023-charge]
+that it is absolutely necessary for this modeling approach to take into account
+the strong ions, which are those assumed to be completely dissociated at any pH.
+In contrast to the physico-chemical model of Stewart[@Stewart1978], the
+charge-balance model (CBM, henceforth) has no separation between independent and
+dependent variables. Under the CBM, all that can be known about the system is
+gathered in the expression of charge in an explicitly stated chemical system.
+Consequently, if, as an example, $\mathrm{pH}$ is known, and charge is a known
+deterministic function of the chemical specification, inverse inferences on
+unknown variables can, at least sometimes, be obtained by root finding, since we
+know that the charge must be zero. In simple systems, such as a single
+monovalent weak acid and strong ions, a single unknown variable is explicitly
+estimable[@Ring2022].
+
+Based on explicit description of the chemistry, an expression of the charge of a
+given solution can always be specified. It is an interesting question if this
+allows _in silico_ modeling of any physiological problem and if every charge
+expression has a physiological interpretation. Another question which remains
+unanswered is how CBM could allow identification of $[\mathrm{H}^{+}]$ at
+several orders below its purported main determinants such as strong ions in the
+presence of unavoidable measurement errors in the determinants much larger than
+$[\mathrm{H}^{+}]$[@Seifter2014].  Actual sensitivity analysis of data generated
+from CBM indicates that say 10% variation in $[\mathrm{H}^{+}]$ has similar
+impact on the calculated charge as 10% variation in SID components or
+$\mathrm{p}\mathrm{CO_2}$, and the success with reproducing the measured
+$\mathrm{pH}$ based on concrete measurements from a large sample of intensive
+care unit patients also supports the notion[@Ring2016]. But still, these results
+are in need for a real explanation. The more mundane question focused here is if
+CBM is rich enough to allow inverse inference in more complex systems such as
+with multiprotic buffers.
+
+This question arose under experiments with HEPES buffered sulfate containing
+fluids[@Ring2023-acid] in which HEPES was modeled as a monovalent buffer with
+pKa $7.5$[@Ring2010] since it was found that estimated $\mathrm{pH}$ was much
+too low when observed $\mathrm{pH}$ was less than about $6$. HEPES is a
+zwitterion with formula 2-[4-(2-Hydroxyethyl)piperazin-1-yl]ethane-1-sulfonic
+acid and structure as shown in \@ref(fig:hepes).
 
@@
 
-Wikipedia informs [@WP:HEPES] that HEPES has two pK values, one at 3 and
-one at 7.5, but the reference given [@Johnson2014] in Wikipedia states
-that "3" comes from the sulfonate group and 7.5 from one of the
-piperazine nitrogens.  At $20^\circ\mathrm{C}$ Good et al [@Good1966]
-has pKa 3 and 7.55 without detailing which groups are at work, while
-another study [@AlGhobashy2014] mentions that 3 and 7.5 both belongs to
-piperazine amines and informs that the sulfonate group is dissociated
-over "almost the whole $\mathrm{pH}$ range".  Furthermore, is has been stated that
-the nitrogen at the $\mathrm{OH}$ end of HEPES is protonated in methanol
-and the nitrogen in the sulfonate end is protonated in water
-[@Sledz2010].
+Wikipedia informs[@WP:HEPES] that HEPES has two $\mathrm{p}K$ values, one at $3$
+and one at $7.5$, but the reference given[@Johnson2014] in Wikipedia states that
+"3" comes from the sulfonate group and $7.5$ from one of the piperazine
+nitrogens. At $20^\circ\mathrm{C}$ Good et al[@Good1966] has pKa $3$ and $7.55$
+without detailing which groups are at work, while another study[@AlGhobashy2014]
+mentions that $3$ and $7.5$ both belong to piperazine amines and informs that
+the sulfonate group is dissociated over "almost the whole $\mathrm{pH}$ range".
+Furthermore, it has been stated that the nitrogen at the $\mathrm{OH}$ end of
+HEPES is protonated in methanol and the nitrogen in the sulfonate end is
+protonated in water[@Sledz2010]. This same reference stated that $\mathrm{p}K_a$
+of the sulfonic group was between $1$ and $3$ but Wikipedia[@WP:Sulfonic_acid]
+has it that sulfonic acids have $\mathrm{p}K_a$ values of $-2.8$ to $-1.9$.
 
-Hence it was not clear how best to model the $\mathrm{pH}$-dependent
-charge on HEPES to estimate $\mathrm{pH}$ based on known chemistry.
-Furthermore, it was desired to make clear if the $\mathrm{p}K$ values
-for HEPES could be elucidated based on measured $\mathrm{pH}$. Herein,
-these questions will be approached by first developing a model for the
+Hence it was not clear how best to model the $\mathrm{pH}$-dependent charge on
+HEPES to estimate $\mathrm{pH}$ based on known chemistry. Furthermore, it was
+desired to make clear if the $\mathrm{p}K$ values for HEPES could be elucidated
+based on measured $\mathrm{pH}$ in fluids with known composition. Normally, the
+$\mathrm{p}K$ values are defined by titration but given the apparent
+disagreement for HEPES it was of interest to see if the CBM offered another
+avenue for understanding.
+
+Herein, these questions will be approached by first developing a model for the
 mean charge on HEPES as a function of $\mathrm{pH}$ and applying this to
-proposed values of set of underlying pKs using data obtained previously
-[@Ring2023-acid]. Next, it will be examined if data generated with
+proposed values of sets of underlying pKs using data obtained
+previously[@Ring2023-acid]. Next, it will be examined if data generated with
 specified $\mathrm{p}K$ values over extended or restricted range of
-$\mathrm{pH}$ values allows recovery of these same $\mathrm{p}K$ values.
-Accepting that $\mathrm{pH} = -1$ would hardly be realized in biological
-fluid and, anyway would require dramatic ionic strengths invalidating
-conventional understanding of $\mathrm{pH}$ modeling, the consequence
-for modeling of assuming the first $\mathrm{p}K_1$ fixed at different
-low values will be examined. In so far as the developed model for
-$\mathrm{pH}$ dependent charge on HEPES implies a strict ordering of
-$\mathrm{p}K$ values, as will be seen in the Methods section, the
-consequence for recovery of known $\mathrm{p}K$ values of assuming
-strict ordering will be assessed. The final model, thus developed, was
-used with the original observed data to obtain best values for the
-underlying $\mathrm{p}K$ values and the fit assessed. Finally, the
-sensitivity of the final model to imposed noise in $\mathrm{SID}$,
-$\mathrm{[HEPES]}$, or $\mathrm{pH}$ will be presented.
+$\mathrm{pH}$ values allows recovery of these same $\mathrm{p}K$ values by
+non-linear regression. Accepting that negative pH values would hardly be
+realized in biological fluid and, anyway, would require dramatic ionic strengths
+making conventional understanding of pH modeling problematic, the consequence
+for modeling of assuming the first $\mathrm{p}K_1$ fixed at different low values
+will be examined. In so far as the developed model for pH dependent charge on
+HEPES implies a strict ordering of pK values, as will be seen in the Methods
+section, it is not clear if the free pK values can be recovered from in silico
+data without such order imposed in the non-linear regression, so this will be
+examined.
+
+To demonstrate how binding is the nature of CBM with the proposed model of
+charge on HEPES, noise free in silico data was next generated with fixed
+$\mathrm{p}K_1$ to examine the effect of pH coverage relative to $\mathrm{p}K_2$
+and $\mathrm{p}K_3$ as used in the simulations.
+
+Finally, normal, Cauchy and uniform noise were imposed on $\mathrm{pH}$ to
+examine the sensitivity of the estimates to noise ratio and noise form, number
+of replications and realized $\mathrm{pH}$. Regarding the last point,
+uncertainty in _measured_ $\mathrm{pH}$ is determined by many independent
+factors with a probably combined standard-deviation of $0.02$ [@Meinrath2000] at
+neutral pH and higher uncertainty at low and high $\mathrm{pH}$ values
+[@Leito2002].
 
@@
 
-```{r echo=FALSE}
-# A dash! A dash! My kingdom for a dash!
-# (The things that rmarkdown users have to do without
-# \usepackage{mhchem})
-```
+<!--
+A dash! A dash! My kingdom for a dash!
+(The things that rmarkdown users have to do without \usepackage{mhchem})
+-->
+
 $\mathrm{I: H^{+}N_1\text-HEP\text-N_4H^{+}SO_3H}$ $\sim \mathrm{p}K_{a1}$ on
@@
 difference[@Ring2022] and $K_w$ the auto-dissociation of water, the new
-charge-balance model for HEPES is:
+CBM model for HEPES is:
 
@@
 enforced by a difference method documented in supplementary materials.
+
+Results
+=======
+
+Observed data consisted of $107$ measurements of $\mathrm{pH}$ from $4.63$ to
+$8.23$ with HEPES from $48.9 \mathrm{mM}$ to $50 \mathrm{mM}$ and SID from
+$–1 \mathrm{mM}$ to $40 \mathrm{mM}$.
+
+```{r predicted-reference, fig.cap = 'Predicted-reference plots for various ways of modelling the pH data', echo = FALSE}
+readRDS('observed_final_250124.rds')
+```
 

### 2024-02-04 observed_final_250124.R

@@ -1,0 +1,265 @@
+# Observed data: 86 instances with variable SID and HEPES and measured pH from
+# the sulfate paper plus 21 with variable SID  fixed HEPES and measured pH.
+# Fitted pH obtained for monovalent HEPES with pKa 7.55 using charge-balance modeling
+# and with trivalent HEPES with pK = c(-1,3,7.55) using the Kildeberg algorithm and
+# (identical) ad hoc formula derived here. Also the formulation of Johnson 2014 with
+# pK1 = 3 and pK2 = 7.55
+# 
+#  First get observed data
+load('HEPES.RData') # DD: 86 lines data.frame SID, HEPES, pH obs
+BUFC = 0.050
+pHobs  <- c(4.63,4.68,4.72,4.77,4.83,4.9,4.96,5.04,5.12,5.21,5.3,
+            5.39,5.48,5.55,5.63,5.69,5.74,5.8,5.85,5.89,5.93)
+
+SIDobs<- c(-seq(10,1)*1e-4,0,seq(1,10)*1e-4)
+SID <- c(DD$SID,SIDobs)
+pH <- c(DD$OBS,pHobs)
+HEPES <- c(DD$HEPES,rep(BUFC,21))
+
+
+
+kw <- 1e-14
+
+
+library(nlsr)
+library(ggplot2)
+library(ggpubr)
+
+HEPESFUNC <- function(H,SID,HEPTOT,pK1,pK2,pK3) {
+  XX <- (H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
+  IV <- HEPTOT/(1+XX)
+  I <- IV*H^3/(10^-pK3*10^-pK2*10^-pK1)
+  II <- IV*H^2/(10^-pK3*10^-pK2)
+  III <- IV*H/10^-pK3
+  H - kw/H + SID + I*2 + II - IV
+}
+
+
+
+pHm <- Vectorize(function(SID, HEPTOT, pK1, pK2, pK3)
+  -log10(uniroot(HEPESFUNC,c(1e-20,1),tol=1e-20,maxiter=1e4,
+                 HEPTOT=HEPTOT,SID = SID, pK1=pK1,pK2=pK2,pK3=pK3)$root))
+
+coefpp <- function(model) {
+  co <- coef(model)
[…]
